---
title: "Generalized Additive Model for Crime Forecasting"
author: "Chatty"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
# Global chunk options: echo = TRUE shows all code in the rendered output
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This notebook fits a **Generalized Additive Model (GAM)** to crime incident data.
The goal is to estimate the probability of a crime occurring at a given
spatial location (grid cell) and time, while accounting for day-of-study trends
and cyclic time-of-day patterns.

The three smooth terms used are:

* `te(grid_longitude, grid_latitude)` — a 2-D tensor-product spline that
  captures spatial crime hotspots.
* `s(day_of_crime)` — a thin-plate spline that captures the overall temporal
  trend across the study period.
* `s(grid_time, bs = "cc")` — a **cyclic cubic spline** that models the
  time-of-day effect so that midnight (hour 0) and the end of the day
  (hour 24) join smoothly.

---

## 1 — Setup

```{r libraries}
# Core data-wrangling and visualisation
library(tidyverse)

# Generalized Additive Models
library(mgcv)
```

---

## 2 — Data

```{r load-or-generate-data}
# ---------------------------------------------------------------------------
# Try to load the real dataset from a local (relative) path.
# If the file is not present — e.g. when running on GitHub Actions — we
# generate a synthetic dataset that has the same structure so that the rest
# of the notebook runs end-to-end without modification.
# ---------------------------------------------------------------------------

data_file <- "crime_prediction_dataset.csv"

if (file.exists(data_file)) {
  # Load the real dataset
  df <- read_csv(data_file, show_col_types = FALSE)
  message("Loaded real dataset: ", nrow(df), " rows.")
} else {
  # ---- Synthetic data generation ----------------------------------------
  # Replace this block with your own data once you place the CSV in the
  # repository root (or adjust `data_file` to point to it).
  message("Real dataset not found. Generating synthetic crime data for demonstration.")

  set.seed(42)
  n_crimes <- 600

  # Partition crimes across three groups:
  #   40 % cluster near (0.25, 0.30) — hotspot A (e.g. a city centre)
  #   40 % cluster near (0.75, 0.70) — hotspot B (e.g. a second district)
  #   20 % random background noise
  n1 <- round(n_crimes * 0.40)
  n2 <- round(n_crimes * 0.40)
  n3 <- n_crimes - n1 - n2

  df <- tibble(
    masked_longitude = c(
      rnorm(n1, mean = 0.25, sd = 0.07),   # hotspot A
      rnorm(n2, mean = 0.75, sd = 0.09),   # hotspot B
      runif(n3, 0, 1)                        # background
    ),
    masked_latitude = c(
      rnorm(n1, mean = 0.30, sd = 0.07),
      rnorm(n2, mean = 0.70, sd = 0.09),
      runif(n3, 0, 1)
    )
  ) %>%
    mutate(
      # Clip coordinates to the valid masked range [0, 1]
      masked_longitude = pmin(pmax(masked_longitude, 0), 1),
      masked_latitude  = pmin(pmax(masked_latitude,  0), 1),

      # Continuous time of day (hours, 0–24)
      time_of_crime = runif(n_crimes, 0, 24),

      # Study day index (1–14, representing two weeks of data)
      day_of_crime  = sample(1:14, n_crimes, replace = TRUE)
    )
  # -----------------------------------------------------------------------
}
```

```{r eda}
# Quick structural overview of the dataset
str(df)
```

---

## 3 — Define the Spatial and Temporal Grid

```{r define-grid}
# ---------------------------------------------------------------------------
# STEP 1: DEFINE THE GRID BOUNDARIES
# ---------------------------------------------------------------------------
# Coordinates are masked to [0, 1].  We discretise them into a 10×10 grid
# (11 break points → 10 bins per axis).
lon_breaks  <- seq(0, 1, length.out = 11)   # 10 longitude bins
lat_breaks  <- seq(0, 1, length.out = 11)   # 10 latitude bins

# Time of day runs 0–24 h; we use 24 one-hour bins.
time_breaks <- seq(0, 24, by = 1)           # 24 time-of-day bins

# Compute the centre coordinate of each bin so the GAM works with
# continuous (rather than integer-index) values.
lon_centers  <- head(lon_breaks,  -1) + diff(lon_breaks)  / 2
lat_centers  <- head(lat_breaks,  -1) + diff(lat_breaks)  / 2
time_centers <- head(time_breaks, -1) + diff(time_breaks) / 2
```

---

## 4 — Map Raw Events to Grid Cells

```{r map-to-grid}
# ---------------------------------------------------------------------------
# STEP 2: MAP RAW DATA TO THE GRID  (grid indices s_i and t_j)
# ---------------------------------------------------------------------------
# Each crime event is assigned to the nearest grid cell (x, y) and time bin.
df_binned <- df %>%
  mutate(
    # Longitude → column index 1..10
    grid_x_index = as.numeric(cut(masked_longitude, breaks = lon_breaks,
                                  include.lowest = TRUE)),
    # Latitude  → row index 1..10
    grid_y_index = as.numeric(cut(masked_latitude,  breaks = lat_breaks,
                                  include.lowest = TRUE)),
    # Hour      → time-bin index 1..24
    time_index   = as.numeric(cut(time_of_crime,    breaks = time_breaks,
                                  include.lowest = TRUE))
  ) %>%
  # Drop rows where a coordinate falls exactly outside the grid boundaries
  drop_na(grid_x_index, grid_y_index, time_index)

# Aggregate: count how many crimes occurred in each (cell, day, hour) combination.
# Mark each observed cell as incident = 1.
actual_crimes <- df_binned %>%
  group_by(grid_x_index, grid_y_index, day_of_crime, time_index) %>%
  summarise(crime_count = n(), .groups = "drop") %>%
  mutate(incident = 1L)
```

---

## 5 — Build the Full Grid with Background Zeros

```{r build-full-grid}
# ---------------------------------------------------------------------------
# STEP 3: CREATE THE BACKGROUND (incident = 0)
# ---------------------------------------------------------------------------
# A GAM with a binary response requires both positive cases (crimes) and
# negative cases (no crime).  We expand every combination of location, day,
# and time and label unobserved cells as incident = 0.

all_days <- unique(df$day_of_crime)

# Cartesian product of all grid cells × days × time bins
full_grid <- expand_grid(
  grid_x_index = seq_along(lon_centers),   # 1..10
  grid_y_index = seq_along(lat_centers),   # 1..10
  day_of_crime = all_days,
  time_index   = seq_along(time_centers)   # 1..24
)

# Left-join actual crimes onto the full grid; missing cells get incident = 0.
model_data <- full_grid %>%
  left_join(actual_crimes,
            by = c("grid_x_index", "grid_y_index", "day_of_crime", "time_index")) %>%
  mutate(
    incident = replace_na(incident, 0L),   # 1 = crime observed, 0 = no crime

    # Replace integer indices with the physical centre coordinates so the
    # GAM smooths over continuous space/time rather than arbitrary integers.
    grid_longitude = lon_centers[grid_x_index],
    grid_latitude  = lat_centers[grid_y_index],
    grid_time      = time_centers[time_index]
  )

cat("model_data dimensions:", nrow(model_data), "rows ×", ncol(model_data), "cols\n")
cat("Prevalence (% cells with a crime):",
    round(100 * mean(model_data$incident), 2), "%\n")
```

---

## 6 — Fit the GAM

```{r fit-gam}
# ---------------------------------------------------------------------------
# STEP 4: FIT THE GENERALIZED ADDITIVE MODEL
# ---------------------------------------------------------------------------
# Response  : incident (binary 0/1)  → family = binomial(logit)
# Predictors:
#   te(grid_longitude, grid_latitude, k=5) — 2-D spatial smooth (hotspot map)
#   s(day_of_crime, k=5)                   — overall temporal trend
#   s(grid_time, bs="cc", k=5)             — cyclic time-of-day effect
# Fitting is done via REML (Restricted Maximum Likelihood) which gives
# well-calibrated smoothing parameters.

gam_model <- gam(
  incident ~ te(grid_longitude, grid_latitude, k = 5) +
             s(day_of_crime, k = 5) +
             s(grid_time, bs = "cc", k = 5),
  family = binomial(link = "logit"),
  data   = model_data,
  method = "REML"
)
```

---

## 7 — Model Summary

```{r model-summary}
# Print the GAM summary:
#   - Parametric coefficients  (intercept)
#   - Approximate significance of smooth terms (edf, F, p-value)
#   - R-sq.(adj), deviance explained, REML score, GCV
summary(gam_model)
```

```{r plot-smooths, fig.width=8, fig.height=4}
# Visualise the estimated smooth effects.
# Each panel shows the partial effect on the log-odds of a crime.
par(mfrow = c(1, 3))
plot(gam_model, pages = 0, shade = TRUE, seWithMean = TRUE,
     main = "GAM Smooth Effects")
par(mfrow = c(1, 1))
```

